Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-08-31 main.Rmd
Finished Section 3

Like most calendars, the Mayan calender used the day as the basic unit of time [@Reingold2001-kf]. The structuring of larger units, weeks, months, years and cycle of years, though, varies substantially between calendars. For example,  the French revolutionary calendar divided each day into 10 "hours", each "hour" into 100 "minutes" and each "minute" into 100 "seconds", the duration of which is 0.864 common seconds. Nevertheless, for any calendar a hierarchy table can be defined. Note that it is not always possible to organize an aperiodic linear granularity in a hierarchy table. Hence, we assume that the hierarchy table consists of periodic linear granularities only, and that the cyclic granularity $C_{G(\ell),G(m)}$ is either circular or quasi-circular.

@@

Methods to obtain multiple-order-up granularity will depend on whether the hierarchy consists of all circular single-order-up granularities or a mix of circular and quasi-circular single-order-up granularities. The methods require the use of integer arithmetic, and hence it is important that the label mapping of the individual single-order-up granularity is an identity function, $L(x) = x$ for all $x$. The label mapping of the resultant multiple-order-up granularity can be chosen depending on the context. Circular single-order-up granularities can be used recursively to obtain a multiple-order-up circular granularity using Equation \ref{eq:eq7} for $\ell < m - 1$ and $P(i, j) = 1, \quad \forall i = j \in \{0, 1, \dots, m-\ell-1\}$.
\begin{equation} \label{eq:eq7}
\begin{split}
C_{G_\ell,G_m}(z) & = C_{G_\ell,G_{\ell+1}}(z) + P(\ell, \ell+1)C_{G_{\ell+1},G_m}(z)\\
  & = C_{G_\ell,G_{\ell+1}}(z) + P(\ell, \ell+1)( C_{G_{\ell+1},G_{\ell+2}}(z) + P(\ell+1, \ell+2)C_{G_{\ell+2}, G_m}(z)) \\
  & = C_{G_\ell,G_{\ell+1}}(z) + P(\ell, \ell+1)C_{G_{\ell+1},G_{\ell+2}}(z) + P(\ell, \ell+1)P(\ell+1, \ell+2)C_{G_{\ell+2}, G_{m}}\\
  & = C_{G_\ell,G_{\ell+1}}(z) + P(\ell, \ell+1)C_{G_{\ell+1},G_{\ell+2}}(z) + P(\ell, \ell+2)C_{G_{\ell+2},G_{\ell+m}}(z)\\
  &\vdots\\
  & = \sum_{i=0}^{m - \ell - 1} P(\ell, \ell+i)C_{G_{\ell+i},G_{\ell+i+1}}(z)\\
\end{split}
\end{equation}
For example, the multiple-order-up granularity $C_{uinal,katun}$ for Mayan calendar could be obtained using Equation \ref{eq:eq8}.
\begin{equation} \label{eq:eq8}
\begin{split}
C_{uinal, baktun}(z)  = & C_{uinal, tun}(z) + P(uinal, tun)C_{tun,katun}(z) + P(uinal, katun)C_{katun, baktun}(z) \\
               = & \lfloor z/20\rfloor \mod 18 + 18*\lfloor z/20*18\rfloor \mod 20  \\
               &~~~ + 18*20\lfloor z/20*18*20\rfloor \mod 20 \\
\end{split}
\end{equation}
We consider the case of only one quasi-circular single order-up granularity in the hierarchy table while computing a multiple-order-up quasi-circular granularity. Any multiple-order-up quasi-circular granularity $C_{\ell, m}(z)$ could then be obtained as a discrete combination of circular and quasi-circular granularities.

Depending on the order of the combination, again two different approaches need to be employed leading to the following cases:

@@

Consider a hierarchy using linear granularities from a Gregorian calendar in \autoref{tab:tab-gregorian}. This is an example of a hierarchy structure which has both circular and quasi-circular single-order-up granularities with day-of-month as the only single-order-up quasi-circular granularity.

Using Equations \ref{eq:multifromsingle-quasi1} and \ref{eq:multifromsingle-quasi2}, we then have:
$$
  C_{hour, month}(z) = C_{hour, day}(z) + P(hour, day)*C_{day, month}(z)
$$
$$
  C_{day, year}(z) = C_{day,month}(z) + \sum_{w=0}^{C_{month, year}(z)-1}(\vert T_{w} \vert),
$$
where $T_w$ is such that $month(w) = \bigcup_{z \in T_w}day(z)$.

@@

@@

Similar to single-to-multiple operations, multiple-to-single operations involve different approaches for all circular single-order-up granularities and a mix of circular and quasi-circular single-order-up granularities in the hierarchy. For a hierarchy table $H_n: (G, C, K)$ with only circular single-order-up granularities in the hierarchy and $\ell_1, \ell_2, m_1, m_2 \in {1, 2, \dots, n}$ and $\ell_2<\ell_1$ and $m_2>m_1$, multiple-order-up granularities could be obtained using \ref{eq:all-circular-multiple}.
\begin{equation} \label{eq:all-circular-multiple}
C_{G_{\ell_1}, G_{m_1}}(z) = \lfloor C_{G_{\ell_2}, G_{m_2}}(z)/P(\ell_2,\ell_1) \rfloor \mod P(\ell_1, m_1)
\end{equation}
For example, in Mayan Calendar, it is possible to compute the single-order-up granularity tun-of-katun from uinal-of-baktun, since $C_{tun, katun}(z) = \lfloor C_{uinal, baktun}(z)/18\rfloor \mod 20$.

@@

Single-order-up quasi-circular granularity could be obtained from multiple-order-up quasi-circular granularity and single/multiple-order-up circular granularity using Equations \ref{eq:multifromsingle-quasi1} and \ref{eq:multifromsingle-quasi2}.

@@

Restructuring time from linear to cyclic time granularities leads to re-organisation of the data structure, where each level of a cyclic granularity corresponds to multiple values of the measured variable. It is common to see summarization of these multiple values through aggregation or an unique summary statistic like mean or median to have each level of cyclic granularity correspond to an unique value of the measured variable. However, this approach hides the distributions of the measured variable induced by the re-organized data structure. Summarizing the distribution of the measured variable using these multiple observations could be a potential way to explore and bring forward different features of the data.

@@

The method is not only restricted to temporal data, and can be generalized to many hierarchical granularities (with continuous and uni-directional nature). We illustrate this with an application to the sport cricket. Although there is no conventional time component in cricket, each ball can be thought to represent an ordering from past to future with the game progressing forward with each ball. In the Twenty20 format, an over will consist of 6 balls (with some exceptions), an innings is restricted to a maximum of 20 overs, a match will consist of 2 innings and a season consists of several matches. Thus, similar to time, there is a hierarchy where ball is nested within overs, overs nested within innings and innings within matches. The idea of cyclic granularities can be likewise mapped to this hierarchy. Example granularites then include ball of the over, over of the innings and ball of the innings. Although most of these cyclic granularities are circular in design of the hierarchy, in application of the rules some granularities are aperiodic. For example, in most cases an over will consist of 6 balls with some exceptions like wide balls or when an innings finishes before the over finishes. Thus, the cyclic granularity ball-of-over will be circular in most cases and aperiodic in others.

@@

# Acknowledgements {-}

>>>>

Like most calendars, the Mayan calendar used the day as the basic unit of time [@Reingold2001-kf]. The structuring of larger units, weeks, months, years and cycle of years, though, varies substantially between calendars. For example,  the French revolutionary calendar divided each day into 10 "hours", each "hour" into 100 "minutes" and each "minute" into 100 "seconds", the duration of which is 0.864 common seconds. Nevertheless, for any calendar a hierarchy table can be defined. Note that it is not always possible to organize an aperiodic linear granularity in a hierarchy table. Hence, we assume that the hierarchy table consists of periodic linear granularities only, and that the cyclic granularity $C_{G(\ell),G(m)}$ is either circular or quasi-circular.

@@

Methods to obtain multiple-order-up granularity will depend on whether the hierarchy consists of all circular single-order-up granularities or a mix of circular and quasi-circular single-order-up granularities. Circular single-order-up granularities can be used recursively to obtain a multiple-order-up circular granularity using
\begin{equation} \label{eq:eq7}
C_{G_\ell,G_m}(z)
  = \sum_{i=0}^{m - \ell - 1} P(\ell, \ell+i)C_{G_{\ell+i},G_{\ell+i+1}}(z),
\end{equation}
where $\ell < m - 1$ and $P(i, i) = 1$ for $i=0,1,\dots,m-\ell-1$, and
$C_{B, G}(z)  = z\mod P(B, G)$ as per Equation \eqref{eq:circular-gran}.

For example, the multiple-order-up granularity $C_{\text{uinal},\text{katun}}$ for the Mayan calendar could be obtained using
\begin{align*}
C_{\text{uinal}, \text{baktun}}(z)  &=  C_{\text{uinal}, \text{tun}}(z) + P(\text{uinal}, \text{tun})C_{\text{tun},\text{katun}}(z) + P(\text{uinal},\text{katun})C_{\text{katun}, \text{baktun}}(z) \\
               &=  \lfloor z/20\rfloor \mod 18 + 18\lfloor 18z/20\rfloor \mod 20
                + 18\times20\lfloor 18\times20\times z/20\rfloor \mod 20.
\end{align*}

Now consider the case where there is one quasi-circular single order-up granularity in the hierarchy table while computing a multiple-order-up quasi-circular granularity. Any multiple-order-up quasi-circular granularity $C_{\ell, m}(z)$ could then be obtained as a discrete combination of circular and quasi-circular granularities.

Depending on the order of the combination, two different approaches need to be employed leading to the following cases:

@@

@@

For example, the Gregorian calendar (\autoref{tab:tab-gregorian}) has day-of-month as a single-order-up quasi-circular granularity, with the other granularities being circular. Using Equations \eqref{eq:multifromsingle-quasi1} and \eqref{eq:multifromsingle-quasi2}, we then have:
$$
  C_{hour, month}(z) = C_{hour, day}(z) + P(hour, day)*C_{day, month}(z)
$$
$$
  C_{day, year}(z) = C_{day,month}(z) + \sum_{w=0}^{C_{month, year}(z)-1}(\vert T_{w} \vert),
$$
where $T_w$ is such that $month(w) = \bigcup_{z \in T_w}day(z)$.

@@

Similar to single-to-multiple operations, multiple-to-single operations involve different approaches for all circular single-order-up granularities and a mix of circular and quasi-circular single-order-up granularities in the hierarchy. For a hierarchy table $H_n: (G, C, K)$ with only circular single-order-up granularities and $\ell_1, \ell_2, m_1, m_2 \in {1, 2, \dots, n}$ and $\ell_2<\ell_1$ and $m_2>m_1$, multiple-order-up granularities can be obtained using \eqref{eq:all-circular-multiple}.
\begin{equation} \label{eq:all-circular-multiple}
C_{G_{\ell_1}, G_{m_1}}(z) = \lfloor C_{G_{\ell_2}, G_{m_2}}(z)/P(\ell_2,\ell_1) \rfloor \mod P(\ell_1, m_1)
\end{equation}
For example, in the Mayan Calendar, it is possible to compute the single-order-up granularity tun-of-katun from uinal-of-baktun, since $C_{tun, katun}(z) = \lfloor C_{uinal, baktun}(z)/18\rfloor \mod 20$.

@@

Single-order-up quasi-circular granularity can be obtained from multiple-order-up quasi-circular granularity and single/multiple-order-up circular granularity using Equations \eqref{eq:multifromsingle-quasi1} and \eqref{eq:multifromsingle-quasi2}.

@@

Restructuring time from linear to cyclic time granularities leads to re-organization of the data structure, where each level of a cyclic granularity corresponds to multiple values of the measured variable. It is common to see summarization of these multiple values through aggregation or an unique summary statistic like mean or median to have each level of cyclic granularity correspond to an unique value of the measured variable. However, this approach hides the distributions of the measured variable induced by the re-organized data structure. Summarizing the distribution of the measured variable using these multiple observations could be a potential way to explore and bring forward different features of the data.

@@

The method is not only restricted to temporal data, and can be generalized to many hierarchical granularities (with continuous and uni-directional nature). We illustrate this with an application to the sport cricket. Although there is no conventional time component in cricket, each ball can be thought to represent an ordering from past to future with the game progressing forward with each ball. In the Twenty20 format, an over will consist of 6 balls (with some exceptions), an innings is restricted to a maximum of 20 overs, a match will consist of 2 innings and a season consists of several matches. Thus, similar to time, there is a hierarchy where ball is nested within overs, overs nested within innings and innings within matches. The idea of cyclic granularities can be likewise mapped to this hierarchy. Example granularities then include ball of the over, over of the innings and ball of the innings. Although most of these cyclic granularities are circular in design of the hierarchy, in application of the rules some granularities are aperiodic. For example, in most cases an over will consist of 6 balls with some exceptions like wide balls or when an innings finishes before the over finishes. Thus, the cyclic granularity ball-of-over will be circular in most cases and aperiodic in others.

@@

# Acknowledgments {-}
